---
title: "Insights on Storm and Bad Weather"
author: "Koushik Nagarajan"
date: "22 June 2015"
output: html_document
---

#Summary
  This insight is looking at the strom and bad weather data from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. 
  This insight is to understand the which event types caused huge damages to humanity and as well economically. 

#Data Loading

Loading the raw data from repdata-data-StormData.csv.bz2
[Link to Download Strom Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

```{r cache=TRUE}
data <- read.csv("data//repdata-data-StormData.csv.bz2")
data$EVTYPE <- tolower(data$EVTYPE)
```


##Data Processing 

* Step 1: Summarise the Event Types based on Fatalities 
* Step 2: Summarise the Event Types based on Injuries
* Step 3: Summarise the Event Types based on Damages

```{r echo=FALSE}

summaryFatalities <- aggregate(data$FATALITIES ~ data$EVTYPE, data = data, FUN = sum)
colnames(summaryFatalities) <- c("EVTYPE", "FATALITIES")
 summaryFatalities <- summaryFatalities[order(summaryFatalities$EVTYPE), ]

summaryInjuries <- aggregate(data$INJURIES ~ data$EVTYPE, data = data, FUN = sum)
colnames(summaryInjuries) <- c("EVTYPE", "INJURIES")
summaryInjuries <- summaryInjuries[order(summaryInjuries$EVTYPE), ]

summaryHarmful <- cbind(summaryFatalities, summaryInjuries$INJURIES)
colnames(summaryHarmful) <- c("EVTYPE", "FATALITIES", "INJURIES")

summaryHarmful <- summaryHarmful[summaryHarmful$FATALITIES > 0, ]
summaryHarmful <- summaryHarmful[order(summaryHarmful$INJURIES, decreasing = TRUE), ]


summaryDamage <- aggregate(data$PROPDMG ~ data$EVTYPE, data = data, FUN = sum)
colnames(summaryDamage) <- c("EVTYPE", "DAMAGES")
```

#Most harmful with respect to population health
```{r}
library(ggplot2)
library(sqldf)
DF=sqldf("SELECT *,FATALITIES+INJURIES AS TotalEffect from summaryHarmful ORDER BY TotalEffect DESC LIMIT 10")
ggplot(data = DF, aes(x =EVTYPE, y = TotalEffect))+
        geom_bar(stat = "identity")+xlab("Event Type")+ylab("Number of Harmful Events")+
        labs(title="Top 10 Event Types With Highest Effect On Humans")

```


#Events have the greatest economic consequences
```{r}
top10Damage=sqldf("SELECT *  from summaryDamage ORDER BY DAMAGES DESC LIMIT 10")
ggplot(data = top10Damage, aes(x =EVTYPE, y = DAMAGES))+
        geom_bar(stat = "identity")+xlab("Event Type")+ylab("Damage Done")+
        labs(title="Top 10 Event Types With Highest Effect On Economy")
```

#Result

* Tornado created the huge effect on humans as per this research.
* Tornado, flash flood and tstm wind caused the huge damages to economy.
